Workshop overview:

-   For loops
-   Conditional (if/else) statements
-   Additional practice problems

# 0. GETTING STARTED

```{r}
library(tidyverse) #data manipulation
library(plyr) #to save loop output in dataframe 

```

We will be using two built in R dataset in this workshop:

1\. the PlantGrowth dataset. It has information about plant weight by three groups (1 control and 2 treatment groups).

2\. the iris dataset, which has information about iris species petal sizes.

```{r}
plants <- PlantGrowth # create a data frame
head(plants) #see the first six rows of the dataframe
```

# 1. FOR LOOPS

For loop's basic structure

for (i in vector/list/matrix/dataframe){ some operation with i }

one step in a loop = one 'iteration'

numeric vector example:

```{r}
#a simple loop: 

numbers <- 1:10 #creating a vector of numbers 1 through 10 

for (i in numbers){
  print(i)
}


```

Add a delay in the loop

```{r}

for (i in numbers){
  print(i)
  Sys.sleep(0.5) #this staggers each output by 0.5 seconds
}
```

In this illustrative example, you need to include print, or the loop will run without producing output.

Why include print? It is helpful to see how many iterations of the loop have been completed to see if

1\. The loop is still going and not stuck.

2\. To calculate the amount of time left for the code to finish running.

```{r}
matrix <- matrix(nrow = 3, ncol = 5, c(1:15))
matrix

for (i in matrix){
  print(i)
}
```

dataframe example:

```{r}
###looping through the unique treatment groups
for (i in unique(plants$group)){
  print(i)
}

```

You can replace i with something more meaningful

```{r}
#note i is an object and never in quotes. i is commonly-used 
#but arbitrary. You can make it more meaningful to you if thats easier, e.g.: 

for (treatment in unique(plants$group)){
  print(treatment)
}
```

Applying an operation by looping through the dataset

```{r}
###finding the maximum weight for the control group
max(plants$weight[plants$group=="ctrl"])

#uses square brackets to subset, will be using this throughout the workshop
```

## PRACTICE 1

Find the max weight for the two treatment groups

```{r}

for(i in unique(plants$group)){
  x <- max(plants$weight[plants$group == i])
  print(i)
  print(x)
}

```

individual components we need to build our for loop:

1.  a list to cycle though: `unique(plants$group)`
2.  an operation inside the loop: `x <- max(plants$weight[plants$group == i])`
3.  for loop structure: `for(i in list){}`

```{r}
#building this into a for loop, for all treatment groups
for (i in unique(plants$group)){
  x <- max(plants$weight[plants$group == i])
  print(x)
}

```

Can save the output in a dataframe

```{r}
## create an empty dataframe
maxdf = NULL

## for loop
for (i in unique(plants$group)){
  
  ## operation from before
  x <- max(plants$weight[plants$group == i])
  
  ## make x into a dataframe to be able to combine it to the existing output dataframe
  x = as.data.frame(x)
  
  ## add a group indicator to x
  x$treatment_group = i
  
  ## add current loop iteration output to exisiting output
  maxdf = rbind.fill(maxdf, x)
  
}
```

## PRACTICE 2

Another example, with a different dataset, `iris`. Find the mean `Petal.Length` for each `Species`

```{r}

meanpl = NULL
species_list = c(unique(iris$Species)) 

for (i in species_list){
  ## find mean
  x = mean(iris$Petal.Length[iris$Species==i])
  ## save output
  x = as.data.frame(x)
  x$sp = i
  meanpl = rbind.fill(meanpl, x)
  
}
```

# 2. IF/ELSE STATEMENTS

Basic structure of an if statement:

if (logical statement){operation true}

can be expanded to an if/else statement:

if (logical statement){ operation if true} else{ operation if false}

examples of logical statements:

x \> 5

x == 4

x == "cat"

x != 0

x %in% c("cat", "dog", "mouse")

Note: an empty else statement "else{}" means "do nothing"

Simple example:

```{r}

x = 6 #set x to 6 

if (x > 5){
  print("x is greater than 5")
}

```

Other example

```{r}
x = 4 #set x to 4 

#add an else statement 
if (x > 5){
  print("x is greater than 5")
} else {
  print("x is less than 5")
}
```

Multiple if statements can be joined together

```{r}
## try different values of a
a=3

if (a > 5) {
  print("a is greater than 5")
} else if (a == 5) {  ## Note the space between "else" and "if" here. 
  print("a = 5")
} else {
  print("a is smaller than 5")
}

```

if/else statement become very useful when embedded within a for loop:

```{r}

#e.g., create a binary variable for control or treatment group in the plants dataset

plants$category <- NA

for (i in 1:nrow(plants)) {
  if (plants$group[i] == 'ctrl') {
    plants$category[i] = 0 ## 0 for control
  } else {
    plants$category[i] = 1 ## 1 for treatment
  }
}

head(plants)
```

Note: for loops will stop running if they produce an error. To keep a for loop going, add tryCatch inside the loop. This is advanced, so we don't go into it, but it may be useful for you as your loops become more advanced <https://cran.r-project.org/web/packages/tryCatchLog/vignettes/tryCatchLog-intro.html>

## PRACTICE 3

For the `iris` datastet, identify if the `Species` **starts with a v or not** in a new column.

```{r}


```

# 4. Tips for how to build loops

Build parts of the loop individually.

-   Check that the loop cycles correctly, this is why we use print!

-   If saving output, make sure that each iteration of the loop have a unique identifier to link the output with each data subset. For example, if your ANOVA is for different metrics, you would want to know what data each ANOVA output links to.

-   Does the code inside the loop work? For example, if you have code for an ANOVA inside the loop, can you use that code to run an ANOVA outside of the loop.

-   To troubleshoot the loop, you can run the lines inside the {} individually.
